The Astropy Project

### 2017-11-16 main.tex
Update on Overleaf.

The development of the \astropypkg core package began as a largely
community-driven effort to standardize core functionality for astronomical
software in \python.
In this way, its genesis differs from but builds upon many substantial and
% * <bsipocz@gmail.com> 2017-11-13T13:13:08.884Z:
%
% how do astroy builds upon these? or it's more like builds upon the developer experience?
%
%
% ^.
former astronomical software development efforts that were commissioned or
initiated through large institutional support, for example IRAF \citep[developed
at NOAO;][]{IRAF}, MIDAS \citep[developed at ESO;][]{MIDAS}, or Starlink
\citep[originally developed by a consortium of UK institutions and now
maintained by the East Asian Observatory;][]{starlink1982,starlink2013}.
% More recently, community-driven efforts have seen significant success in the astronomical sciences \citep{yt}
\inlinecomment{BMS}{I like the idea to reference yt here}
%
\python\footnote{\url{https://www.python.org/}} is an increasingly popular,
general-purpose scripting language that is available under a permissive open
source software license free of charge for all major operating systems.
Stable and well-developed packages provide support for array arithmetic
\citep[\package{numpy};][]{numpy}, a wide variety of functions for scientific
computing \citep[\package{scipy};][]{numpy}, and publication-quality plotting
\citep[\package{matplotlib};][]{matplotlib}. Tens of thousands of other packages
are available which can help with tasks that are not astronomy specific but
might be performed in the course of astronomical research, for example interfacing with databases or statistical inference.

@@

As of version 2.0, \astropypkg contains $212244$ lines of code\footnote{This
line count includes comments, as these are often as important for
maintainability as the code itself.  Without comments there are $142197$ lines
of code.} contributed by $232$ unique contributors over $19270$ \texttt{git}
commits.
The commit distribution follows an approximate power-law (with log-slope of
$\approx -0.47$) \inlinecomment{AG}{round to -0.5?  This is astronomy, after all.}, meaning most of the commits were and are contributed by a
relatively small fraction of the contributor pool.
Roughly half of the commits were contributed by the top $4$
contributors, while $\sim 90$ percent of the commits came from the top $24$
contributors.

@@

Central to the success of \astropy is an open environment where anybody can
contribute to the project.
However, this model leads to ``organic'' growth where different features are
implemented by different people with different programming styles and
interfaces.
Thus, \astropy has a mechanism to more formally propose significant changes to
the core package (e.g., the release cycle; \citealt{ape2})\inlinecomment{AG}{Is the release cycle really part of the core package?  That's how this reads.  Is there something more concretely linked to the code we can use as an example here?}, to plan out major
new features, or institute new organization-wide policies (e.g.,
code of conduct; \citealt{ape8}).
This mechanism is called ``Astropy Proposal for Enhancement'' (APE) and are
modeled after the ``Python Enhancement Proposals'' (PEP) that guide the
development of the \python programming language.
In an APE, one or more authors describe in detail the proposed changes or
additions, including a rationale for the changes, how these changes will be
implemented, and, in the case of code, what the interface will be \citep{ape1}.
The APEs are discussed and refined by the community before much work is invested
into a detailed implementation; anyone is welcome to contribute to these
discussions during the open consideration period by leaving comments on the
respective pull request.\footnote{\url{https://github.com/astropy/astropy-APEs}}
In previous APEs a community consensus emerged and APEs are accepted and become
the basis for future work at this point.
In cases where consensus cannot be reached, the
\astropy coordination committee can decide to close the discussion and
make an executive decision based on the community input on the APE in question.

@@

The \astropypkg core package is available from several of the \python distributions for scientific
% * <mattwcraig@gmail.com> 2017-10-27T07:13:51.093Z:
%
% > \astropypkg i
% I changed a few of the macros in here to refer to the package rather than the project because it is the package being discussed here.
%
% ^.
computing, such as Anaconda or Enthought Canopy, which include also some of
\astropy's affiliated packages.
Alternatively, \astropypkg can be readily installed from the Python Package
Index\footnote{\url{https://pypi.python.org/pypi}} (PyPI) using pip, the
recommended command-line tool for installing \python packages
that is included by default since \python 2 $\geq$ 2.7.9 and \python 3 $\geq$ 3.4.
For most recent \astropypkg releases, pre-built wheel distributions are obtainable
from PyPI for Windows, Mac OS X and Linux systems that provide faster
installation compared to the source distribution.
% hamogu: I think that's too detailed for a journal paper.
%     It's in the docs if anyone needs it.
%Optionally, the version of
%the package and the dependencies needed to run or build the software are given
%by a specifier to the pip package manager as described in PEP 440\footnote{For
%more detailed installation instructions and optional flags see
%\url{http://docs.astropy.org/en/stable/install.html\#installing-astropy}}.

Packaged versions of \astropy are also available for several
operating systems that are managed by a system's package manager,
including Debian, Ubuntu, Fedora, Gentoo, Arch Linux, Mac OS X (e.g.,
via brew, macports or fink) and FreeBSD\@.  However, on occasion these packages
may be outdated compared with the latest stable release of \astropy accessible
from the PyPI\@. Pre-built binary conda packages for all of the affiliated packages are also available  via the \texttt{astropy} conda channel for Linux, Max OS X, and MS Windows.
% * <mattwcraig@gmail.com> 2017-10-27T07:16:27.310Z:
%
% > re-built binary conda packages
% Feel free to remove if this is too detailed. I do think the available of the whole ecosystem across all platforms has helped speed its adoption.
%
% ^.

@@

% \subsection{Analytic Functions}
% moved to models and depreciated --
% just here for completeness at the moment
\inlinecomment{BMS}{I suggest to order the subsection lexicographically as reading the draft the current order seemed very random (unlike at the docs page where they the grouping logic is obvious)}

@@

The \texttt{astropy.units} subpackage adds support for representing units and
numbers with associated units --- ``quantities'' --- in code.
Historically, quantities in code have often been represented simply as numbers,
with units implied or noted via comments in the code because of considerations
about speed: having units associated with numbers inherently adds overhead to
numerical operations.
In \texttt{astropy.units}, \texttt{Quantity} objects subclass the \texttt{numpy}
array object and have been designed with speed in mind.
As of \astropypkg version 2.0, units and quantities are prevalent in most other
\astropypkg subpackages and are thus a key concept for using the package as a
whole.
Units are intimately entwined in the definition of astronomical coordinates and
thus nearly all functionality in the \texttt{astropy.coordinates} subpackage
(see \sectionname~\ref{sec:coordinates}) depends on this functionality.
For most other subpackages, quantities are at least accepted, and often expected
by default.

@@

\subsubsection{Speed improvements} The \texttt{Quantity} object is now significantly
        faster than in early versions of \astropypkg. This is in large part due
        to the fact that \texttt{Quantity} subclasses the \texttt{numpy.ndarray}
        object, but even relative to more recent versions of \astropypkg,
        operations with \texttt{Quantity} objects are generally faster. This is
        largely thanks to improvements in the way \texttt{numpy} arrays can be
        subclassed.

@@

@@

The \texttt{astropy.constants} subpackage provides a selection of physical and
astronomical constants as \texttt{Quantity} objects (see
\sectionname~\ref{sec:units}).
A brief description of this package was given in \citep{astropy}.
In version 2.0, the built-in constants have been organized into modules for
specific versions of the constant values.
For example, physical constants have \texttt{codata2014} \citep{codata2014} and
\texttt{codata2010} versions.
Astronomical constants are organized into \texttt{iau2015} and \texttt{iau2012}
modules to indicate their sources (resolutions from the International
Astronomical Union, IAU).
The \texttt{codata2014} and \texttt{iau2015} versions are combined into the
default constant value version: \texttt{astropyconst20}.
For compatibility with \astropypkg version 1.3 and earlier, the \texttt{astropyconst13}
namespace is still available and provides access to the adopted versions of the
constants from earlier versions of \astropypkg.
To use previous versions of the constants as \emph{units} (e.g., solar masses),
the values from the desired namespace have to be imported directly; with version
2.0, \texttt{astropy.units} uses the \texttt{astropyconst20} versions.

@@

\subsubsection{Derivatives of coordinate representations}
    As mentioned above, the \texttt{Representation} classes act like
    three-dimensional vectors.
    \texttt{astropy.coordinates} now also supports handling first derivatives of
    vectors / representations through the new \texttt{Differential} classes.
    The \texttt{Differential} classes are currently used internally within the
    \texttt{Frame} classes to store the velocity data.

@@

\inlinecomment{TR}{This is a first draft and I need to dig into the details again to finish some of the text, and I also need to make the table/figure}.

@@

The \package{astropy.time} subpackage focuses on supporting time scales (e.g.,
UTC, TAI, UT1) and time formats (e.g., Julian date, modified Julian date) that
are commonly used in astronomy.
This functionality is needed, for example, to calculate barycentric corrections
or sidereal times.
\package{astropy.time} is currently built on the ERFA (\citealt{erfa}) C
library, which replicates the Standards of Fundamental Astronomy (SOFA;
\citealt{sofa}) but is licensed under a three-clause BSD license.
The package was described in detail in previous work \citep{astropy} and has
stayed stable for the last several versions of \astropypkg.
Thus, in what follows, we only highlight significant changes or new features
since the previous \astropy paper.

@@

@@

@@

@@

>>>>

\inlinecomment{Tom R}{the 'this community' above is a bit too general - do we mean it is the only community in astronomy, or in the Python astronomy world? I'd change this to either 'to become this community for Python astronomy software' or 'to become such a community'. For instance yt had a community before we came along for the simulation community, so I want to make sure we don't claim we are the only community out there.}

The development of the \astropypkg core package began as a largely
community-driven effort to standardize core functionality for astronomical
software in \python.
In this way, its genesis differs from but builds upon many substantial and
% * <bsipocz@gmail.com> 2017-11-13T13:13:08.884Z:
%
% how do astroy builds upon these? or it's more like builds upon the developer experience?
%
%
% ^.
former astronomical software development efforts that were commissioned or
initiated through large institutional support, for example IRAF \citep[developed
at NOAO;][]{IRAF}, MIDAS \citep[developed at ESO;][]{MIDAS}, or Starlink
\citep[originally developed by a consortium of UK institutions and now
maintained by the East Asian Observatory;][]{starlink1982,starlink2013}.
% More recently, community-driven efforts have seen significant success in the astronomical sciences \citep{yt}
\inlinecomment{BMS}{I like the idea to reference yt here}
%
\inlinecomment{Tom R}{Python calls itself a programming language or general-purpose language, not specifically scripting language. I would remove 'scripting' below.}

\python\footnote{\url{https://www.python.org/}} is an increasingly popular,
general-purpose scripting language that is available under a permissive open
source software license free of charge for all major operating systems.
Stable and well-developed packages provide support for array representation and arithmetic
\citep[\package{numpy};][]{numpy}, a wide variety of functions for scientific
computing \citep[\package{scipy};][]{numpy}, and publication-quality plotting
\citep[\package{matplotlib};][]{matplotlib}. Tens of thousands of other packages
are available which can help with tasks that are not astronomy specific but
might be performed in the course of astronomical research, for example interfacing with databases or statistical inference.

@@

As of version 2.0, \astropypkg contains $212244$ lines of code\footnote{This
line count includes comments, as these are often as important for
maintainability as the code itself.  Without comments there are $142197$ lines
of code.} contributed by $232$ unique contributors over $19270$ \texttt{git}
commits.
The commit distribution \inlinecomment{Tom R}{too vague, maybe 'the distribution of number of commits as a function of ...'} follows an approximate power-law (with log-slope of
$\approx -0.47$) \inlinecomment{AG}{round to -0.5?  This is astronomy, after all.}, meaning most of the commits were and are contributed by a
relatively small fraction of the contributor pool.
Roughly half of the commits were contributed by the top $4$
contributors, while $\sim 90$ percent of the commits came from the top $24$
contributors.

@@

Central to the success of \astropy is an open environment where anybody can
contribute to the project.
However, this model leads to ``organic'' growth where different features are
implemented by different people with different programming styles and
interfaces.
Thus, \astropy has a mechanism to more formally propose significant changes to
the core package (e.g., the release cycle; \citealt{ape2})\inlinecomment{AG}{Is the release cycle really part of the core package?  That's how this reads.  Is there something more concretely linked to the code we can use as an example here?}, to plan out major
new features, or institute new organization-wide policies (e.g.,
code of conduct; \citealt{ape8}).
This mechanism is called ``Astropy Proposal for Enhancement'' (APE) and are
modeled after the ``Python Enhancement Proposals'' (PEP) that guide the
development of the \python programming language.
In an APE, one or more authors describe in detail the proposed changes or
additions, including a rationale for the changes, how these changes will be
implemented, and, in the case of code, what the interface will be \citep{ape1}.
The APEs are discussed and refined by the community before much work is invested
into a detailed implementation; anyone is welcome to contribute to these
discussions during the open consideration period by leaving comments on the
respective pull request.\footnote{\url{https://github.com/astropy/astropy-APEs}} \inlinecomment{Tom R}{first mention of a pull request in the context of an APE - need to make it clear that APEs are proposed by PRs}
In previous APEs a community consensus emerged and APEs are accepted and become
the basis for future work at this point.
In cases where consensus cannot be reached, the
\astropy coordination committee can decide to close the discussion and
make an executive decision based on the community input on the APE in question.

@@

The \astropypkg core package is available from several of the \python distributions for scientific
% * <mattwcraig@gmail.com> 2017-10-27T07:13:51.093Z:
%
% > \astropypkg i
% I changed a few of the macros in here to refer to the package rather than the project because it is the package being discussed here.
%
% ^.
computing, such as Anaconda or Enthought Canopy, which include also some of
\astropy's affiliated packages.
Alternatively, \astropypkg can be readily installed from the Python Package
Index\footnote{\url{https://pypi.python.org/pypi}} (PyPI) 
\inlinecomment{Tom R}{make footnote link directly to astropy on PyPI?}
using pip, the
recommended command-line tool for installing \python packages
that is included by default since \python 2 $\geq$ 2.7.9 and \python 3 $\geq$ 3.4.

\inlinecomment{Tom R}{is this actually true? I installed Python 3.6 from source recently and it didn't have the pip command readily available. What it does have is the ensurepip module, but we might want to be careful how we phrase this here. Maybe just remove mention of pip being included by default?}
For most recent \astropypkg releases, pre-built wheel distributions are obtainable
from PyPI for Windows, Mac OS X and Linux systems that provide faster
installation compared to the source distribution.
% hamogu: I think that's too detailed for a journal paper.
%     It's in the docs if anyone needs it.
%Optionally, the version of
%the package and the dependencies needed to run or build the software are given
%by a specifier to the pip package manager as described in PEP 440\footnote{For
%more detailed installation instructions and optional flags see
%\url{http://docs.astropy.org/en/stable/install.html\#installing-astropy}}.

Packaged versions of \astropy are also available for several
operating systems that are managed by a system's package manager,
including Debian, Ubuntu, Fedora, Gentoo, Arch Linux, Mac OS X (e.g.,
via brew, macports or fink) and FreeBSD\@.  However, on occasion these packages
may be outdated compared with the latest stable release of \astropy accessible
from the PyPI\@. Pre-built binary conda packages for all of the affiliated packages are also available  via the \texttt{astropy} conda channel for Linux, Mac OS X, and MS Windows.
% * <mattwcraig@gmail.com> 2017-10-27T07:16:27.310Z:
%
% > re-built binary conda packages
% Feel free to remove if this is too detailed. I do think the available of the whole ecosystem across all platforms has helped speed its adoption.
%
% ^.

@@

% \subsection{Analytic Functions}
% moved to models and depreciated --
% just here for completeness at the moment
\inlinecomment{BMS}{I suggest to order the subsection lexicographically as reading the draft the current order seemed very random (unlike at the docs page where they the grouping logic is obvious)}
\inlinecomment{Tom R}{the only issue is if we rely on having e.g. units described first if they are used later}

@@

The \texttt{astropy.units} subpackage adds support for representing units and
numbers with associated units --- ``quantities'' --- in code.
Historically, quantities in code have often been represented simply as numbers,
with units implied or noted via comments in the code because of considerations
about speed: having units associated with numbers inherently adds overhead to
numerical operations.
In \texttt{astropy.units}, \texttt{Quantity} objects subclass the \texttt{numpy}
array object and have been designed with speed in mind.

\inlinecomment{Tom R}{I think saying 'subclass' here is too detailed and is mentioned below in 3.1.1 if needed - 'make it possible to attach units to numpy arrays' or something like that?}

As of \astropypkg version 2.0, units and quantities are prevalent in most other
\astropypkg subpackages and are thus a key concept for using the package as a
whole.
Units are intimately entwined in the definition of astronomical coordinates and
thus nearly all functionality in the \texttt{astropy.coordinates} subpackage
(see \sectionname~\ref{sec:coordinates}) depends on this functionality.
For most other subpackages, quantities are at least accepted, and often expected
by default.

@@

\subsubsection{Speed improvements}

\inlinecomment{Tom R}{I wonder if we might just want to mention at the top level that performance has been improved in places? This seems a bit too detailed}

The \texttt{Quantity} object is now significantly
        faster than in early versions of \astropypkg. This is in large part due
        to the fact that \texttt{Quantity} subclasses the \texttt{numpy.ndarray}
        object, but even relative to more recent versions of \astropypkg,
        operations with \texttt{Quantity} objects are generally faster. This is
        largely thanks to improvements in the way \texttt{numpy} arrays can be
        subclassed.

@@

\inlinecomment{Tom R}{this is a new feature of table, not units, so I think we don't want to mention this here}

@@

The \texttt{astropy.constants} subpackage provides a selection of physical and
astronomical constants as \texttt{Quantity} objects (see
\sectionname~\ref{sec:units}).
A brief description of this package was given in \citep{astropy}.
In version 2.0, the built-in constants have been organized into modules for
specific versions of the constant values.
For example, physical constants have \texttt{codata2014} \citep{codata2014} and
\texttt{codata2010} versions.
Astronomical constants are organized into \texttt{iau2015} and \texttt{iau2012}
modules to indicate their sources (resolutions from the International
Astronomical Union, IAU).
The \texttt{codata2014} and \texttt{iau2015} versions are combined into the
default constant value version: \texttt{astropyconst20}.
For compatibility with \astropypkg version 1.3 and earlier, the \texttt{astropyconst13}
namespace \inlinecomment{Tom R}{jargon alert: 'namespace'} is still available and provides access to the adopted versions of the
constants from earlier versions of \astropypkg.
To use previous versions of the constants as \emph{units} (e.g., solar masses),
the values from the desired namespace have to be imported directly; with version
2.0, \texttt{astropy.units} uses the \texttt{astropyconst20} versions.

@@

\subsubsection{Derivatives of coordinate representations}

\inlinecomment{Tom R}{does this really warrant a separate section or is it just a byproduct of the velocity stuff that can be mentioned in 3.3.2?}

    As mentioned above, the \texttt{Representation} classes act like
    three-dimensional vectors.
    \texttt{astropy.coordinates} now also supports handling first derivatives of
    vectors / representations through the new \texttt{Differential} classes.
    The \texttt{Differential} classes are currently used internally within the
    \texttt{Frame} classes to store the velocity data.

@@

@@

The \package{astropy.time} subpackage focuses on supporting time scales (e.g.,
UTC, TAI, UT1) and time formats (e.g., Julian date, modified Julian date) that
are commonly used in astronomy.
This functionality is needed, for example, to calculate barycentric corrections
or sidereal times.
\package{astropy.time} is currently built on the ERFA (\citealt{erfa}) C
library, which replicates the Standards of Fundamental Astronomy (SOFA;
\citealt{sofa}) but is licensed under a three-clause BSD license.
The package was described in detail in  \citet{astropy} and has
stayed stable for the last several versions of \astropypkg.
Thus, in what follows, we only highlight significant changes or new features
since the previous \astropy paper.

@@

\inlinecomment{Tom R}{I would remove this tiny section. As for units, I think we should only mention the whole mixin stuff in the table section, not for every single object that can go in a table}

@@

\inlinecomment{Tom R}{this is waaaaay too short and jargony. It's a very cool feature that deserves more reader-friendly text!}

@@

\inlinecomment{Tom R}{is a sub-subpackage a thing, or is it just 'subpackage' all the way down?}

@@

\inlinecomment{Tom R}{too detailed? I don't think we need to mention every single performance enhancement, right?}
